---
title: "Effect of VLCAD mutation - network diagram"
author: "Mark Ziemann"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    fig_width: 7
    fig_height: 7
theme: cosmo
---

Source: TBA

## Introduction

Here we are repeating the network diagrams.
The purpose of this is to cut down the analysis just so we can run the minimal packages for making the network diagrams.
This is important to prepare these charts to be included in the mitch package.

```{r,packages}

load("dge.Rdata")

suppressPackageStartupMessages({
  library("reshape2")
  library("dplyr")
  library("mitch")
  library("network")
})

knitr::opts_chunk$set(dev = 'svg') # set output device to svg

```

## Network diagrams

In these charts the darkness of the colour indicates the enrichment score.
The size of the node indicates the number of genes in the set driving the enrichment.
The width of the edges indicates the proportion of shared genes.

```{r,netfuncs}

map2color <- function(x,pal,limits=NULL){
  if (is.null(limits)) limits=range(x)
  pal[findInterval(x,seq(limits[1],limits[2],length.out=length(pal)+1), all.inside=TRUE)]
}

gs2net <- function(gset,em,colfunc=colorRampPalette(c("blue", "white","red"))(n=100)){
  gset <- gset[order(names(gset))]
  mydf <- dplyr::bind_rows(lapply(gset, as.data.frame.list))
  rownames(mydf) <- names(gset)
  j <- apply(mydf,1,function(x) {
    apply(mydf,1,function(y) {
      length(intersect(x,y) ) / length(union(x,y))
    })
  })
  j[lower.tri(j)] <- NA
  j[lower.tri(j,diag=TRUE)] <- 0
  jl <- reshape2::melt(j)
  jl <- jl[which(jl$Var1 != jl$Var2),]
  jl <- jl[which(jl$value != 1),]
  jl <- jl[order(-jl$value),]
  jl <- head(jl,length(gset)*2)
  jl$edgeSize = jl$value/sum(jl$value)
  nodes <- sort(union(jl[,1],jl[,2]))
  lengths <- unlist(lapply(gset,length))
  lengths <- lengths[names(lengths) %in% nodes]
  nodes <- data.frame("nodes"=nodes,"lengths"=lengths)
  nodes$vertexsize <- sqrt(nodes$lengths/sum(nodes$lengths) * 100)
  nodes$es <- em[match(nodes$nodes,em$set),"s.dist"]
  nodes$colours <- map2color(nodes$es,colfunc)
  jl2 <- apply(jl[,1:2],2,as.character)
  jlnet <- network(jl2)
  jlnet$val <- lapply(jlnet$val, function(x) {
    vn <- x[[2]]
    vn <- substr(vn, 1, 60)
    if (nchar(vn) == 60 ) {
      vn <- paste(vn,"...",sep="")
    }
  x[[2]] <- vn
  return(x)
  } )
  plot(jlnet, displaylabels = TRUE, label.col = "steelblue",
       edge.lwd = c(jl$edgeSize) * 100,
       arrowhead.cex = 0,
       label.cex = 0.8, vertex.border = "white",vertex.cex = nodes$vertexsize,
       vertex.col = nodes$colours, edge.col = "black")
  E1=min(nodes$es)
  E5=max(nodes$es)
  E3=mean(c(E1,E5))
  EE <- c(E1,E3,E5)
  legcols <- map2color(EE,colfunc)
  legend("topleft", legend=signif(EE,2) ,title="ES",box.lty=0,
    fill=legcols, cex=0.8)
  S1=min(nodes$vertexsize)
  FRAG=S1/10
  S5=max(nodes$vertexsize)
  S3=mean(c(S1,S5))
  SS=c(S1-FRAG,0,S5-FRAG)
  L1=min(nodes$lengths)
  L5=max(nodes$lengths)
  LL=paste(" ",c(L1,"",L5))
  legend("topright", legend=LL ,title="no. genes",box.lty=0,
    pt.cex=SS*1, cex=0.9 , pch=19,col="darkgray")
  J1=min(jl$edgeSize)
  FRAG=J1*3
  J5=max(jl$edgeSize)
  J3=mean(c(J1,J5))
  JJ=c(J1,J3,J5)
  JL=JJ+FRAG
  legend("bottomleft", legend=signif(JJ,2) , lwd=JL*50, title="Jaccard",
    box.lty=0, cex=0.9 , lty=1, col="black")
}

netplot <- function(eres,FDR=0.05,n_sets=20) {
  scores <- eres[[1]][,1]
  names(scores) <- rownames(eres[[1]])
  gs=eres[[2]]
  eres <-eres[[4]]
  up <- head(subset(eres,p.adjustANOVA<FDR & s.dist > 0),n_sets)
  up_gs <- up[,1]
  up_gs <- gs[which(names(gs) %in% up_gs)]

  dn <- head(subset(eres,p.adjustANOVA<FDR & s.dist < 0),n_sets)
  dn_gs <- dn[,1]
  dn_gs <- gs[which(names(gs) %in% dn_gs)]
  topgs_up <- lapply(1:length(up_gs),function(i) {
    gsname <- names(up_gs)[i]
    genes <- up_gs[[i]]
    gene_scores <- scores[which(names(scores) %in% genes)]
    top_genes <- names(which(gene_scores>2))
    return(top_genes)
  })
  names(topgs_up) <- names(up_gs)
  topgs_dn <- lapply(1:length(dn_gs),function(i) {
    gsname <- names(dn_gs)[i]
    genes <- dn_gs[[i]]
    gene_scores <- scores[which(names(scores) %in% genes)]
    top_genes <- names(which(gene_scores<2))
    return(top_genes)
  })
  names(topgs_dn) <- names(dn_gs)
  gs2net(gset=topgs_up,em=eres,colfunc=colorRampPalette(c("pink","darkred"))(n=100))
  gs2net(gset=topgs_dn,em=eres,colfunc=colorRampPalette(c("darkblue","lightblue"))(n=100))
}

```

GO BP Network.

```{r,gobp_net,fig.width=14,fig.height=6}

netplot(eres=mres_gobp,FDR=0.05,n_sets=20)

```

## Session information

```{r,save}

sessionInfo()

```
